---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

## Quick start

Welcome to the `exprso` GitHub page! Let's get started.

```{r, eval = FALSE}
library(devtools)
devtools::install_github("tpq/exprso")
library(exprso)
```

```{r, echo = FALSE, message = FALSE}
library(exprso)
set.seed(1)
```

## Importing data

To import data, we use the `exprso` function. This function has two arguments.

```{r}
data(iris)
array <- exprso(iris[1:80, 1:4], iris[1:80, 5])
```

## Pre-processing data

Functions with a `mod` prefix pre-process the data.

```{r}
array <- modTransform(array)
array <- modNormalize(array, c(1, 2))
```

## Split data

Functions with a `split` prefix split the data into training and test sets.

```{r}
arrays <- splitSample(array, percent.include = 67)
array.train <- arrays$array.train
array.test <- arrays$array.valid
```

## Select features

Functions with a `fs` prefix select features.

```{r}
array.train <- fsStats(array.train, top = 0, how = "t.test")
```

## Build models

Functions with a `build` prefix build models.

```{r}
mach <- buildSVM(array.train,
                 top = 50,
                 kernel = "linear",
                 cost = 1)
pred <- predict(mach, array.train)
pred <- predict(mach, array.test)
```

```{r, eval = FALSE}
calcStats(pred)
```

## Deploy pipelines

Functions with a `pl` prefix deploy high-throughput learning pipelines.

```{r, results = "hide"}
pl <- plGrid(array.train,
             array.test,
             how = "buildSVM",
             top = c(2, 4),
             kernel = "linear",
             cost = 10^(-3:3),
             fold = NULL)
```

```{r}
pl
```

Read the exprso vignettes for more details.
